---
title: "Extracting Census Tract FIPS codes from geocoordinates"
output: html_notebook
author: Taylor Alarcon
---
# Setting up working environment and data cleaning

The datasets used for this example:

- [MappingPoliceViolence.org]("https://mappingpoliceviolence.org/)
- [US Census Demographic and Economic Data for Census Tracts and Counties]("https://www.kaggle.com/datasets/muonneutrino/us-census-demographic-data/data) from Kaggle

**Motivation**: This is a data wrangling and API-calling exercise as part of a broader research project looking at the impact that instances of local police violence have on the neighborhoods in which that violence occurs. From MappingPoliceViolence.org I have a geocoded repository of instances of fatal police interactions nationwide, but unfortunately the "tract" variable is quite messy in the original dataset. For that reason I am leveraging the spatial package *sf* and the census API-calling package "tidycensus* to manually impute FIPS codes for Census Tracts that saw instances of police violence. For the purposes of this tutorial, I will only be imputing FIPS codes for instances in California.

### Load libraries
```{r}
#load libraries
library(tidyverse)
library(haven) # for reading dta files
library(lubridate) # for datetime
library(tigris) # for census data
library(sf) # for spatial data
library(tidycensus) # census api
```

### Load datasets
```{r}
# Mapping Police Violence data
mpv_raw <- read_csv("/Users/taylor/Desktop/phd/research-local/data/police-violence-data/MPV03282023.csv")

# US Census Tract Demographics dataset from Kaggle
census_tract_demographics_raw <- read.csv("/Users/taylor/Desktop/phd/research-local/data/housing-neighborhoods/us-census-tracts/kaggle census tract demographic data/acs2017_census_tract_data.csv")
```

Note: Exploring different parts of the data before wrangling.
```{r}
View(census_tract_demographics_raw)
dim(census_tract_demographics_raw)
colnames(census_tract_demographics_raw)
```


```{r}
View(mpv_raw)
mpv_raw$tract # missing a lot of census tracts
sum(is.na(mpv_raw$tract))
count(mpv_raw$tract)
dim(mpv_raw)
```

### Clean up MPV data to be up to 2017, since the Census data only goes up to 2017 
```{r}
mpv_raw$date <- as.Date(mpv_raw$date, format = "%m/%d/%Y")
mpv_raw$date
```


```{r}
mpv_2017 <- mpv_raw %>%
  filter(date < "2018-01-01")
```


```{r}
# check sanity
dim(mpv_2017) # 5405 instances of fatal police violence nationwide up to 2017
min(mpv_2017$date) # 2012-04-07
max(mpv_2017$date) # sanity check
```


# Getting the FIPS codes (Census Tract IDs) for California instances of fatal police violence up to 2017 

Finally a solution that works: calling the tidycensus api to get FIPS codes 
Link: https://stackoverflow.com/questions/52248394/get-census-tract-from-lat-lon-using-tigris

### Filtering MPV data to just California

```{r}
# filter to just California
mpv_2017_CA <- mpv_2017 %>%
  filter(state == "CA")
```


```{r}
mpv_2017_CA # 896 instances in CA
```


```{r}
# getting just latitude and longitude for instances in CA
mpv_2017_CA_long_lat_only <- mpv_2017_CA %>%
  select(longitude, latitude)
```

### Loading CA shapefile from tidycensus API + point data geoprocessing
```{r}
# getting CA census tract data (shapefile) from the Census API, per tutorial link above
ca <- tidycensus::get_acs(state = "CA", geography = "tract",
              variables = "B19013_001", geometry = TRUE)
```


```{r}
# convert to sf object of points, so that we can match it to the shapefile above
mpv_2017_CA_long_lat_sf <- mpv_2017_CA_long_lat_only %>%
  filter(!is.na(latitude), !is.na(longitude)) %>% # get rid of NAs
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(ca)) # converting lat and long numbers to geometry object
```

### Visualization(s)
```{r}
# plot the sf object created above now that it's been geoprocessed
plot(mpv_2017_CA_long_lat_sf)
```


```{r}
# better visualization of the point data, overlaying CA shapefile 
ggplot(ca) +
  geom_sf() +
  geom_sf(data = mpv_2017_CA_long_lat_sf) +
  ggtitle("Mapping Police Violence in California 2012-2017 
          (896 observations)")
```

### Extracting the FIPS codes for Census Tracts where police violence occurred
```{r}
# variable for all CA census tracts -- retrieving from tidycensus api
ca_tracts <- tracts("CA", class = "sf") %>%
  select(GEOID, TRACTCE)
```


```{r}
# creating boundaries around all CA census tracts
CA_bbox <- st_bbox(ca_tracts)
```


```{r}
# just st_join to find intersections of point data and/within CA census tracts
CA_points_tract <- st_join(mpv_2017_CA_long_lat_sf, ca_tracts) # spits out the tracts associated with the points 
CA_points_tract # 953 observations since many Census Tract IDs are duplicates (multiple instances of police violence)
```

### Create the dataframe to store all the fips codes for CA
```{r}
tract_IDs_CA <- c(CA_points_tract$GEOID)
```


```{r}
length(tract_IDs_CA) # 953 CA instances with census tracts
length(unique(tract_IDs_CA)) # 863 unique census tracts 
```

# Final list of Census Tract IDs

### Remove duplicates for final list
```{r}
tract_IDs_CA_final <- c(unique(tract_IDs_CA))
```

```{r}
length(tract_IDs_CA_final) # sanity check to make sure there's only 863
print(tract_IDs_CA_final) # print Census Tract FIPS codes that experienced police violence through 2017
```


